---
title: "Data Structure Presentation"
output: html_document
---

This file is used to generate plots:

- 0. log2CountHeatmap.pdf: Heatmap of matched and filtered counts on log2 scale.

- 1. OTU_hist.pdf: Histograms of log2-transformed WTS and 16S read counts.

- 2. median_filter.pdf, median_match.pdf: Scatter plots comparing the log2-transformed WTS and 16S read counts' mean or median for various genera in both filtered and matched data.

- 3. prevalence_WTS.pdf, prevalence_16S.pdf: This showcases genera in 16S and WTS with a prevalence exceeding 10% and highlights the 65 common genera.

- 4. relative_abundance_raw.pdf, relative_abundance_filter.pdf: Relative abundance of top 10 genus stacked bar plots on raw data and filtered data.

- 5. rich_read_raw.pdf, rich_read_filter.pdf: scatterplot between total reads and richness on raw and filtered data.
     rich_read_comp.pdf: scatter plot between 16S richness and WTS richness, and between 16S reads, and WTS reads for both matched and filtered data.

- 6. WTS_16S_top10_venn_diagram.pdf: venn diagram of top 10 genus with the highest abundance in both 16S and WTS filter data.

- 7. Concordance Correlation Coefficient (CCC) between genus-median,  CCC between sample richness, CCC between sample total reads,  number of overlapped genera in the corresponding top10 genera.


```{r import libraries, message=FALSE}
library(ggplot2)
library(ggpubr)
library(dplyr)
library(tidyverse)
library(microViz)
library(patchwork)
library(ggforce)
library(jjb)
library(phyloseq)
library(VennDiagram)
library(scales)
library(knitr)
library(DescTools)
library(circlize)
library(ComplexHeatmap)
library(ggsci)
```


```{r load cleaned data}
M_16S_raw = read.csv("M_16S_raw.csv", row.names = 1)
M_WTS_raw = read.csv("M_WTS_raw.csv", row.names = 1)

M_16S_matched = read.csv("M_16S_matched.csv", row.names = 1)
M_WTS_matched = read.csv("M_WTS_matched.csv", row.names = 1)

M_16S_filter = read.csv("M_16S_filtered.csv", row.names = 1)
M_WTS_filter = read.csv("M_WTS_filtered.csv", row.names = 1)

Raw_M_16S = readRDS("Raw_M_16S.rds")
Raw_M_WTS = readRDS("Raw_M_WTS.rds")

Matched_M_16S = readRDS("Matched_M_16S.rds")
Matched_M_WTS = readRDS("Matched_M_WTS.rds")

Filtered_M_16S = readRDS("Filtered_M_16S.rds")
Filtered_M_WTS = readRDS("Filtered_M_WTS.rds")

TSS_M_16S = readRDS("TSS_M_16S.rds")
TSS_M_WTS = readRDS("TSS_M_WTS.rds")

DRP_M_16S = readRDS("DRP_M_16S.rds")
DRP_M_WTS = readRDS("DRP_M_WTS.rds")

DV_M_16S = readRDS("DV_M_16S.rds")
DV_M_WTS = readRDS("DV_M_WTS.rds")

TMM_M_16S = readRDS("TMM_M_16S.rds")
TMM_M_WTS = readRDS("TMM_M_WTS.rds")

CSS_M_16S = readRDS("CSS_M_16S.rds")
CSS_M_WTS = readRDS("CSS_M_WTS.rds")

GMPR_M_16S = readRDS("GMPR_M_16S.rds")
GMPR_M_WTS = readRDS("GMPR_M_WTS.rds")

Wrench_M_16S = readRDS("Wrench_M_16S.rds")
Wrench_M_WTS = readRDS("Wrench_M_WTS.rds")
```

## 0. Heatmap of matched and filtered samples
```{r, heatmaps of log2counts}
col_fun <- colorRamp2(c(0,14), c("turquoise", "orangered"))

# Create the heatmap with row and column splits, color scale, and specified data range
heat_matched <- Heatmap(cbind(log2(M_16S_matched+1), log2(M_WTS_matched+1)),
        name = "log2(count+1)",
        column_split = rep(c("16S Matched","WTS Matched"), c(ncol(M_16S_matched), ncol(M_WTS_matched))),  # Split columns
        col = col_fun,             # Specify color scale
        show_row_names = FALSE,     # Show row names
        show_column_names = FALSE,   # Show column names
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        row_names_gp = gpar(fontsize = 3),
        heatmap_legend_param = list(legend_direction = "horizontal")
)
heat_filtered <- Heatmap(cbind(log2(M_16S_filter+1), log2(M_WTS_filter+1)),
        name = "log2(count+1)",
        column_split = rep(c("16S Filtered","WTS Filtered"), c(ncol(M_16S_filter), ncol(M_WTS_filter))),  # Split columns
        col = col_fun,             # Specify color scale
        show_row_names = FALSE,     # Show row names
        show_column_names = FALSE,   # Show column names
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        row_names_gp = gpar(fontsize = 3),
        heatmap_legend_param = list(legend_direction = "horizontal")
)
pdf("../Figures/Heatmap_log2Count_Matched.pdf", width = 8, height = 8) 
draw(heat_matched,  heatmap_legend_side = "bottom") 
dev.off()

pdf("../Figures/Heatmap_log2Count_Filtered.pdf", width = 8, height = 8) 
draw(heat_filtered,  heatmap_legend_side = "bottom") 
dev.off()
```


## 1. Histograms of log2-transformed WTS and 16S read counts

```{r OTU_hist.pdf}
p_wts <- ggplot(data = data.frame(WTS = log2(unlist(c(M_WTS_raw))+1)), aes(x = WTS))+
  geom_histogram(aes(y = after_stat(count / sum(count))))+
  labs( x = "log2(count+1)", y = "WTS Proportions") +
  ylim(0,1)+
  theme_classic()

p_16s <- ggplot(data = data.frame(S16 = log2(unlist(c(M_16S_raw))+1)), aes(x = S16))+
  geom_histogram(aes(y = after_stat(count / sum(count))))+
  labs( x = "log2(count+1)", y = "16S Proportions") +
  ylim(0,1)+
  theme_classic()


p_hist <- ggarrange(p_16s, p_wts, nrow=2)

ks.test(log2(unlist(c(M_WTS_raw))+1), log2(unlist(c(M_16S_raw))+1))
ks.test(log2(unlist(c(M_WTS_matched))+1), log2(unlist(c(M_16S_matched))+1))
ks.test(log2(unlist(c(M_WTS_filter))+1), log2(unlist(c(M_16S_filter))+1))


# save OTU_hist.pdf
ggsave(file = "../Figures/OTU_hist.pdf", p_hist, width = 4, height=8)
p_hist
mean((unlist(c(M_WTS_raw))) == 0)
mean((unlist(c(M_16S_raw))) == 0)
summary(unlist(c(M_WTS_raw)))
summary(unlist(c(M_16S_raw)))
```

## 2. Scatter plots comparing the log-transformed WTS and 16S read counts' mean or median for various genera in filtered data/matched data.

```{r median_filter.pdf}
# compute the mean and median of log2 read counts for each genus in both 16S and WTS data

mean_16_filter <- apply(log2(M_16S_filter+1), 1, mean)
mean_WTS_filter <- apply(log2(M_WTS_filter+1), 1, mean)

median_16_filter <- apply(log2(M_16S_filter+1), 1, median)
median_WTS_filter <- apply(log2(M_WTS_filter+1), 1, median)

mean(names(mean_16_filter) == names(mean_WTS_filter))
mean(names(median_16_filter) == names(median_WTS_filter))
DescTools::CCC(mean_16_filter, mean_WTS_filter)$rho.c["est"]
DescTools::CCC(median_16_filter, median_WTS_filter)$rho.c["est"]

med_filter <- data.frame(M_WTS = median_WTS_filter, M_16S = median_16_filter)
mean_filter <- data.frame(M_WTS = mean_WTS_filter, M_16S = mean_16_filter)

# generate scatter plot comparing the filtered log-transformed WTS and 16S read counts' median.
p_filter_median <- ggplot(med_filter, aes(x = M_16S, y = M_WTS, color = "black")) + 
  geom_point() +
  scale_color_manual(values=c( "black")) +
  labs(title = "", x = "16S genus median on log2 scale", y = "WTS genus median on log2 scale") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed")+ 
  theme(plot.title = element_text(hjust = 0.5),legend.position = "none")+
  theme_classic()+
  theme(legend.position = "none")

# generate scatter plot comparing the filtered log2-transformed WTS and 16S read counts' mean.
p_filter_mean <- ggplot(mean_filter, aes(x = M_16S, y = M_WTS, color = "black")) + 
  geom_point() +
  scale_color_manual(values=c( "black")) +
  labs(title = "", x = "16S genus mean on log2 scale", y = "WTS genus mean on log2 scale") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed")+ 
  theme(plot.title = element_text(hjust = 0.5),legend.position = "none") +
  theme_classic()+
  theme(legend.position = "none")

# Save median_filter.pdf
ggsave(p_filter_median, 
       filename="../Figures/median_filter.pdf",
       height=4, width=4)
ggsave(p_filter_mean, 
       filename="../Figures/mean_filter.pdf",
       height=4, width=4)
p_filter_median
p_filter_mean
sum(med_filter$M_16S == 0)
sum(med_filter$M_WTS[med_filter$M_16S == 0]!=0)
```

```{r median_match.pdf}
# compute the mean and median of log2 read counts for each genus in both 16S and WTS data
mean_16_match <- apply(log2(M_16S_matched+1), 1, mean)
mean_WTS_match <- apply(log2(M_WTS_matched+1), 1, mean)

median_16_match <- apply(log2(M_16S_matched+1), 1, median)
median_WTS_match <- apply(log2(M_WTS_matched+1), 1, median)

med_match <- data.frame(M_WTS = median_WTS_match, M_16S = median_16_match)
mean_match <- data.frame(M_WTS = mean_WTS_match, M_16S = mean_16_match)

mean(names(mean_16_match) == names(mean_WTS_match))
mean(names(median_16_match) == names(median_16_match))
DescTools::CCC(mean_16_match, mean_WTS_match)$rho.c["est"]
DescTools::CCC(median_16_match, median_WTS_match)$rho.c["est"]


# generate scatter plot comparing the matched log-transformed WTS and 16S read counts' median.
p_match_median <- ggplot(med_match, aes(x = M_16S, y = M_WTS, color = "black")) + 
  geom_point() +
  scale_color_manual(values=c( "black")) +
  labs(title = "", x = "16S genus median on log2 scale", y = "WTS genus median on log2 scale") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed")+ 
  theme(plot.title = element_text(hjust = 0.5),legend.position = "none")+
  theme_classic()+
  theme(legend.position = "none")

# generate scatter plot comparing the matched log2-transformed WTS and 16S read counts' mean.
p_match_mean <- ggplot(mean_match, aes(x = M_16S, y = M_WTS, color = "black")) + 
  geom_point() +
  scale_color_manual(values=c( "black")) +
  labs(title = "", x = "16S genus mean on log2 scale", y = "WTS genus mean on log2 scale") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed")+ 
  theme(plot.title = element_text(hjust = 0.5),legend.position = "none") +
  theme_classic()+
  theme(legend.position = "none")

ggsave(p_match_median, 
       filename="../Figures/median_match.pdf",
       height=4, width=4)
ggsave(p_match_mean, 
       filename="../Figures/mean_match.pdf",
       height=4, width=4)
ggsave(ggarrange(p_match_mean, p_match_median, nrow = 1), 
       filename="../Figures/mean_median_match.pdf",
       height=4, width=8)
p_match_median
p_match_mean
sum(med_match$M_16S == 0)
sum(med_match$M_WTS[med_match$M_16S == 0]!=0)
```

## 3. Prevalence of 16S and WTS

```{r prevalence_16S.pdf}
# 16S
# calculate prevalence proportion of each genus in 16S raw
prevalence_16S <- apply(M_16S_raw > 0, 1, mean) %>% data.frame()
colnames(prevalence_16S) <- "proportion"
prevalence_16S <-
  prevalence_16S %>% 
  rownames_to_column(var = "genus") %>% 
  arrange(desc(proportion))

# annotate the genus whose prevalence_16S is smaller than 10%
prevalence_16S$Feature <- "remained"
prevalence_16S$Feature[!(prevalence_16S$genus %in% rownames(M_16S_filter))] <- "filtered"
prevalence_16S$Feature <- as.factor(prevalence_16S$Feature)

mean(prevalence_16S$proportion < 0.1)

# after arranging 16S genus from large to small, mark the genera with prevalence higher than 10%, and draw the boundary line at 10%
p_prevalence_16S <- prevalence_16S %>% 
  ggplot(aes(x = reorder(genus,-proportion), y = proportion, color = Feature)) + 
  geom_bar(stat = "identity", alpha=0.5) +
  scale_color_manual(values=c("#999999","black")) +
  geom_hline(yintercept = 0.1, color="red", linetype="dashed", linewidth = 0.5)+
  annotate(geom = "text", x = 150, y = 0.1, label = "0.1",vjust = -1, color = "red", size = 4) + 
  theme(axis.text.x = element_blank(),panel.background = element_rect(fill = "white")) +
  labs(title = "", x = "genus of 16S", y = "prevalence of 16S genus", color = "")

# save prevalence_16S.png
ggsave(filename="../Figures/prevalence_16S.pdf",
       plot=p_prevalence_16S,
       height=4, width=9)
p_prevalence_16S
```

```{r prevalence_WTS.pdf}
# WTS
# identify the prevalence of each genus in WTS
prevalence_WTS <- apply(M_WTS_raw > 0, 1, mean) %>% data.frame()
colnames(prevalence_WTS) <- "proportion"
prevalence_WTS <- prevalence_WTS %>% filter(proportion != 0) %>% rownames_to_column(var = "feature")
prevalence_WTS$Feature <- "filtered"
prevalence_WTS$Feature[prevalence_WTS$feature %in% rownames(M_WTS_filter)] <- "remained"
prevalence_WTS$Feature <- as.factor(prevalence_WTS$Feature)

mean(prevalence_WTS$proportion < 0.1)

# after arranging WTS genus from large to small, mark the genera with prevalence_16S higher than 10%, and draw the boundary line at 10%
p_prevalence_WTS <- prevalence_WTS %>% 
  ggplot(aes(x = reorder(feature,-proportion), y = proportion, color = Feature)) + 
  geom_bar(stat = "identity", alpha=0.5) +
  scale_color_manual(values=c(  "#999999","black")) +
  geom_hline(yintercept = 0.1, color="red", linetype="dashed", size = 0.5)+
  annotate(geom = "text", x = 1500, y = 0.1, label = "0.1",vjust = -1, color = "red", size = 4) + 
  theme(legend.position = "bottom")  +
  theme(axis.text.x = element_blank()) +
  labs(title = "", x = "genus of WTS", y = "prevalence of WTS genus", color = "")
# Save prevalence_WTS.png
ggsave(filename="../Figures/prevalence_WTS.pdf",
       plot=p_prevalence_WTS,
       height=4, width=9)

ggsave(filename = "../Figures/prevalence.pdf", ggarrange(p_prevalence_16S, p_prevalence_WTS, nrow = 1, common.legend = TRUE, legend = "top"), height = 4, width = 16)
ggsave(filename = "../Figures/prevalence_horizontal.pdf", ggarrange(p_prevalence_16S, p_prevalence_WTS, nrow = 2, common.legend = TRUE, legend = "bottom"), height = 8, width = 8)
```


## 4. Genus relative abundance top 10

```{r relative_abundance_raw.pdf}
# 1. identify the top 10 genus with the highest abundance in both 16S and WTS raw data.
toptax_16S <- tax_top(Raw_M_16S, n=10, by=sum)
toptax_WTS <- tax_top(Raw_M_WTS, n=10, by=sum)
tax_com <- toptax_WTS[toptax_WTS %in% toptax_16S] 

# assign unique color codes for each genus and take 16S's colours as standard, coloring shared genus with it.
col_16S <- distinct_palette(n=10, add="grey90")
col_WTS <- distinct_palette(n=20, add="grey90")[11:21]

for (ft in tax_com){
  col_WTS[1:10][toptax_WTS==ft] <-  col_16S[1:10][toptax_16S==ft]
}
names(col_16S) <- c(toptax_16S,"others")
names(col_WTS) <- c(toptax_WTS,"others")
# 2. generate barplots showing the top 10 genera of WTS and 16S raw data
rel_16S <-
  Raw_M_16S %>% comp_barplot(
    tax_level = "Genus",
    n_taxa = 10,
    other_name = "other",
    palette = col_16S,
    merge_other = FALSE
  ) + 
  facet_grid(
    cols = vars (diagnosis),
    scales = "free", space = "free") + 
  #coord_flip() +
  labs(x = NULL) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  theme(legend.position = "bottom", 
        legend.text = element_text(size=10),
        legend.key.size = unit(0.2, 'cm'),
        legend.title = element_text(size=10))+
  ggtitle('16S') + 
  theme(plot.title=element_text(size=12, hjust=0.5))+
  labs(y="Relative abundance")

rel_WTS <-
  Raw_M_WTS %>% comp_barplot(
    tax_level = "Genus",
    n_taxa = 10,
    other_name = "other",
    palette = col_WTS,
    merge_other = FALSE
  ) + 
  facet_grid(
    cols = vars (diagnosis),
    scales = "free", space = "free") + 
  #coord_flip() +
  labs(x = NULL) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  theme(legend.position = "bottom", 
        legend.text = element_text(size=10),
        legend.key.size = unit(0.2, 'cm'),
        legend.title = element_text(size=10)) +
  ggtitle('WTS') +
  theme(plot.title=element_text(size=12, hjust=0.5))+
  labs(y="Relative abundance")

ggsave(filename="../Figures/relative_abundance_raw_horizontal.pdf", rel_16S + rel_WTS + plot_layout(ncol = 2),
       height=8, width=18)
rel_16S + rel_WTS
```


```{r relative_abundance_filter.pdf}
# 1. identify the top 10 genus with the highest expression in both 16S and WTS filter data.
toptax_16S_filter <- tax_top(Filtered_M_16S, n=10, by=sum)
toptax_WTS_filter <- tax_top(Filtered_M_WTS, n=10, by=sum)
tax_com_filter <- intersect(toptax_16S_filter, toptax_WTS_filter)

# assign unique color codes for each genus and take 16S's colours as standard, coloring shared genus with it.
col_16S_filter <- distinct_palette(n=10, add="grey90", pal='kelly')
col_WTS_filter <- distinct_palette(n=20, add="grey90", pal='kelly')[11:21]
for (ft in tax_com_filter){
  col_WTS_filter[1:10][toptax_WTS_filter==ft] <-  col_16S_filter[1:10][toptax_16S_filter==ft]
}

names(col_16S_filter) <- c(toptax_16S_filter,"others")
names(col_WTS_filter) <- c(toptax_WTS_filter,"others")

for (ft in  unique(c(toptax_16S, toptax_WTS))){
  if(ft %in% toptax_WTS_filter){
     col_WTS_filter[1:10][toptax_WTS_filter==ft] <-   ifelse(ft %in% toptax_16S, col_16S[1:10][toptax_16S==ft], col_WTS[1:10][toptax_WTS==ft])
  }
  if(ft %in% toptax_16S_filter){
     col_16S_filter[1:10][toptax_16S_filter==ft]  <-   ifelse(ft %in% toptax_16S, col_16S[1:10][toptax_16S==ft], col_WTS[1:10][toptax_WTS==ft])
  }
}


# 2. generate barplots showing the top 10 genera of WTS and 16S filter data
rel_16S_filter <-
  Filtered_M_16S %>% comp_barplot(
    tax_level = "Genus",
    n_taxa = 10,
    other_name = "other",
    palette = col_16S_filter,
    merge_other = FALSE
  ) + 
  facet_grid(
    cols = vars (diagnosis),
    scales = "free", space = "free") + 
  #coord_flip() +
  labs(x = NULL) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  theme(legend.position = "bottom", 
        legend.text = element_text(size=10),
        legend.key.size = unit(0.2, 'cm'),
        legend.title = element_text(size=10))+
  ggtitle('16S') +
  theme(plot.title=element_text(size=12, hjust=0.5))+
  labs(y="Relative abundance")

rel_WTS_filter <-
  Filtered_M_WTS %>% comp_barplot(
    tax_level = "Genus",
    n_taxa = 10,
    other_name = "other",
    palette = col_WTS_filter,
    merge_other = FALSE
  ) + 
  facet_grid(
    cols = vars (diagnosis),
    scales = "free", space = "free") + 
  #coord_flip() +
  labs(x = NULL) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  theme(legend.position = "bottom", 
        legend.text = element_text(size=10),
        legend.key.size = unit(0.2, 'cm'),
        legend.title = element_text(size=10)) +
  ggtitle('WTS') +
  theme(plot.title=element_text(size=12, hjust=0.5))+
  labs(y="Relative abundance")

ggsave(filename="../Figures/relative_abundance_filter_horizontal.pdf", 
       rel_16S_filter + rel_WTS_filter + plot_layout(ncol = 2),
       height=8, width=18)
```


## 5. Observed richness and total reads

```{r rich_read_raw.pdf}
richness_16S_raw <- ggplot(data = data.frame("total_reads" =  phyloseq::sample_sums(Raw_M_16S),
                         "observed" = phyloseq::estimate_richness(Raw_M_16S, measures = "Observed")[, 1]),
       aes(x = (total_reads), y = (observed))) +
  geom_point() +
  geom_smooth(method="loess", se = FALSE, span = 0.5) +
  labs(x = "16S total reads", y = "16S observed richness") +
  theme(plot.title=element_text(size=9, hjust=0.5)) +
  theme_bw()+
  scale_x_continuous(labels = scales::scientific_format())


richness_WTS_raw <- ggplot(data = data.frame("total_reads" =  phyloseq::sample_sums(Raw_M_WTS),
                         "observed" = phyloseq::estimate_richness(Raw_M_WTS, measures = "Observed")[, 1]),
       aes(x = total_reads, y = (observed))) +
  geom_point() +
  geom_smooth(method="loess", se = FALSE, span = 0.5) +
  labs(x = "WTS total reads", y = "WTS observed richness") +
  theme(plot.title=element_text(size=9, hjust=0.5)) +
  theme_bw()+
  scale_x_continuous(labels = scales::scientific_format())

ggsave(filename="../Figures/rich_read_raw.pdf",ggarrange(richness_16S_raw, richness_WTS_raw, nrow=2),
       height=8, width=4)


M16S_totalreads <- data.frame("total_reads" =  phyloseq::sample_sums(Raw_M_16S),
                         "observed" = phyloseq::estimate_richness(Raw_M_16S, measures = "Observed")[, 1])
MWTS_totalreads <-  data.frame("total_reads" =  phyloseq::sample_sums(Raw_M_WTS),
                         "observed" = phyloseq::estimate_richness(Raw_M_WTS, measures = "Observed")[, 1])
mean(rownames(M16S_totalreads)==rownames(MWTS_totalreads))

plot(M16S_totalreads$total_reads, MWTS_totalreads$total_reads)
abline(a=0,b=1)

plot(M16S_totalreads$observed, MWTS_totalreads$observed)
abline(a=0,b=1)
```

```{r rich_read_match.pdf}
richness_16S_match <- ggplot(data = data.frame("total_reads" =  phyloseq::sample_sums(Matched_M_16S),
                         "observed" = phyloseq::estimate_richness(Matched_M_16S, measures = "Observed")[, 1]),
       aes(x = (total_reads), y = (observed))) +
  geom_point() +
  geom_smooth(method="loess", se = FALSE, span = 0.5) +
  labs(x = "16S total reads", y = "16S observed richness") +
  theme(plot.title=element_text(size=9, hjust=0.5)) +
  theme_bw()+
  scale_x_continuous(labels = scales::scientific_format())


richness_WTS_match <- ggplot(data = data.frame("total_reads" =  phyloseq::sample_sums(Matched_M_WTS),
                         "observed" = phyloseq::estimate_richness(Matched_M_WTS, measures = "Observed")[, 1]),
       aes(x = total_reads, y = (observed))) +
  geom_point() +
  geom_smooth(method="loess", se = FALSE, span = 0.5) +
  labs(x = "WTS total reads", y = "WTS observed richness") +
  theme(plot.title=element_text(size=9, hjust=0.5)) +
  theme_bw()+
  scale_x_continuous(labels = scales::scientific_format())

ggsave(filename="../Figures/rich_read_match.pdf",ggarrange(richness_16S_match, richness_WTS_match, nrow=2),
       height=8, width=4)
```



```{r rich_read_filter.pdf}

richness_16S_filter <- ggplot(data = data.frame("total_reads" =  phyloseq::sample_sums(Filtered_M_16S),
                         "observed" = phyloseq::estimate_richness(Filtered_M_16S, measures = "Observed")[, 1]),
       aes(x = total_reads, y = (observed))) +
  geom_point() +
  geom_smooth(method="loess", se = FALSE, span=0.5) +
  labs(x = "16S total reads", y = "16S observed richness") +
  theme(plot.title=element_text(size=9, hjust=0.5)) +
  theme_bw()+
  scale_x_continuous(labels = scales::scientific_format())


richness_WTS_filter <- ggplot(data = data.frame("total_reads" =  phyloseq::sample_sums(Filtered_M_WTS),
                         "observed" = phyloseq::estimate_richness(Filtered_M_WTS, measures = "Observed")[, 1]),
       aes(x = total_reads, y = (observed))) +
  geom_point() +
  geom_smooth(method="loess", se = FALSE) +
  labs(x = "WTS total reads", y = "WTS observed richness") +
  theme(plot.title=element_text(size=9, hjust=0.5)) +
  theme_bw()+
  scale_x_continuous(labels = scales::scientific_format())

ggsave(filename="../Figures/rich_read_filter.pdf",
       ggarrange(richness_16S_filter,richness_WTS_filter, nrow=2),
       height=8, width=4)



M16SFilter_totalreads <- data.frame("total_reads" =  phyloseq::sample_sums(Filtered_M_16S),
                         "observed" = phyloseq::estimate_richness(Filtered_M_16S, measures = "Observed")[, 1])
MWTSFilter_totalreads <-  data.frame("total_reads" =  phyloseq::sample_sums(Filtered_M_WTS),
                         "observed" = phyloseq::estimate_richness(Filtered_M_WTS, measures = "Observed")[, 1])
mean(rownames(M16S_totalreads)==rownames(MWTS_totalreads))


plot(M16SFilter_totalreads$total_reads, MWTSFilter_totalreads$total_reads)
abline(a=0,b=1)

plot(M16SFilter_totalreads$observed, MWTSFilter_totalreads$observed)
abline(a=0,b=1)
```

```{r rich_match_filter_comp.pdf}
rich_match <- data.frame("M16S" = phyloseq::estimate_richness(Matched_M_16S, measures = "Observed")[, 1],
                         "MWTS" = phyloseq::estimate_richness(Raw_M_WTS, measures = "Observed")[, 1],
                         type = "Matched")
rich_filter <- data.frame("M16S" = phyloseq::estimate_richness(Filtered_M_16S, measures = "Observed")[, 1],
                         "MWTS" = phyloseq::estimate_richness(Filtered_M_WTS, measures = "Observed")[, 1],
                         type = "Filtered")
rich_comp <- ggplot(rbind(rich_match, rich_filter))+
  geom_point(aes(x = M16S, y = MWTS, color = type))+
  geom_abline(intercept = 0, slope = 1)+
  scale_color_bmj()+
  labs(x = "16S observed richness", y = "WTS observed richness", color = "") +
  theme_bw()+
  theme(legend.position = "bottom")


total_match <- data.frame("M16S" = phyloseq::sample_sums(Matched_M_16S),
                         "MWTS" = phyloseq::sample_sums(Matched_M_WTS),
                         type = "Matched")
total_filter <- data.frame("M16S" = phyloseq::sample_sums(Filtered_M_16S),
                         "MWTS" = phyloseq::sample_sums(Filtered_M_WTS),
                         type = "Filtered")
total_comp <- ggplot(rbind(total_match, total_filter))+
  geom_point(aes(x = M16S, y = MWTS, color = type))+
  geom_abline(intercept = 0, slope = 1)+
  scale_color_bmj()+
  labs(x = "16S total reads", y = "WTS total reads", color = "") +
  theme_bw()+
  theme(legend.position = "bottom")

ggsave(filename="../Figures/rich_read_comp.pdf",
       ggarrange(rich_comp, total_comp, nrow=2, common.legend = TRUE, legend = "bottom"),
       height=8, width=4)

```



## 6. venn diagram of top 10 genus with the highest abundance in both 16S and WTS raw and filtered data.

```{r venn_raw.pdf}
top_10_data_16S <- Raw_M_16S@otu_table[rownames(Raw_M_16S@otu_table) %in% toptax_16S, ]
genus_mean_proportions_16S <- rowSums(top_10_data_16S)[toptax_16S]

top_10_data_WTS <- Raw_M_WTS@otu_table[rownames(Raw_M_WTS@otu_table) %in% toptax_WTS, ]
genus_mean_proportions_WTS <- rowSums(top_10_data_WTS)[toptax_WTS]

intersection <- intersect(names(genus_mean_proportions_16S), names(genus_mean_proportions_WTS))
set_16S_only <- setdiff(names(genus_mean_proportions_16S), names(genus_mean_proportions_WTS))  
set_WTS_only <- setdiff(names(genus_mean_proportions_WTS), names(genus_mean_proportions_16S))  

print(paste("Shared genus:", paste(intersection, collapse = ", ")))
print(paste("16S only:", paste(set_16S_only, collapse = ", ")))
print(paste("WTS only:", paste(set_WTS_only, collapse = ", ")))

set_16S <- names(genus_mean_proportions_16S)
set_WTS <- names(genus_mean_proportions_WTS)

venn_data <- list(
  Set1 = set_16S,
  Set2 = set_WTS
)

venn.plot <- venn.diagram(
  x = venn_data,
  category.names = c("16S", "WTS"),   
  filename = NULL,                     
  output = TRUE,                      
  fill = c("skyblue", "pink"),        
  alpha = 0.5,                        
  cex = 1.5,                           
  cat.cex = 1.5,                        
  cat.pos = c(-10, 10),                 
  cat.dist = c(0.05, 0.05),             
  lwd = 2,                             
  fontfamily = "sans",                 
  label.col = "transparent",           
  cat.col = c("blue", "red"),
  scaled = FALSE            
)

values <- genus_mean_proportions_WTS[set_WTS_only]
sorted_set_WTS_only <- set_WTS_only[order(values, decreasing = TRUE)]
rankings <- rank(-genus_mean_proportions_WTS)
sorted_set_WTS_only <- sapply(sorted_set_WTS_only, function(genus) {
  paste(genus, " (", rankings[genus],")",sep="")
})

values <- genus_mean_proportions_16S[set_16S_only]
sorted_set_16S_only <- set_16S_only[order(values, decreasing = TRUE)]
rankings <- rank(-genus_mean_proportions_16S)
sorted_set_16S_only <- sapply(sorted_set_16S_only, function(genus) {
  paste("(", rankings[genus],") ",genus,sep="")
})

values <- genus_mean_proportions_16S[setdiff(names(genus_mean_proportions_16S), set_16S_only)]
intersection_with_values <- intersection[order(values, decreasing = TRUE)]
rankings_16S <- rank(-genus_mean_proportions_16S)
rankings_WTS <- rank(-genus_mean_proportions_WTS)
intersection_with_values <- sapply(intersection_with_values, function(genus) {
  paste("(", rankings_16S[genus],") ", genus, " (", rankings_WTS[genus],")",sep="")
})

pdf("../Figures/WTS_16S_top10_venn_diagram_raw.pdf", width = 9, height = 6)
grid.newpage()
grid.draw(venn.plot)

grid.text(paste(paste(intersection_with_values, collapse = "\n"), collapse = ", "), x = 0.5, y = 0.475, gp = gpar(fontsize = 12.5))
grid.text(paste(paste(sorted_set_16S_only, collapse = "\n"), collapse = ", "), x = 0.325, y = 0.475, gp = gpar(fontsize = 12.5), just = "right") 
grid.text(paste(paste(sorted_set_WTS_only, collapse = "\n"), collapse = ", "), x = 0.675, y = 0.475, gp = gpar(fontsize = 12.5), just = "left")
dev.off()
```

```{r venn_filtered.pdf}
top_10_data_16S <- Filtered_M_16S@otu_table[rownames(Filtered_M_16S@otu_table) %in% toptax_16S_filter, ]
genus_mean_proportions_16S <- rowSums(top_10_data_16S)[toptax_16S_filter]


top_10_data_WTS <- Filtered_M_WTS@otu_table[rownames(Filtered_M_WTS@otu_table) %in% toptax_WTS_filter, ]
genus_mean_proportions_WTS <- rowSums(top_10_data_WTS)[toptax_WTS_filter]

intersection <- intersect(names(genus_mean_proportions_16S), names(genus_mean_proportions_WTS))
set_16S_only <- setdiff(names(genus_mean_proportions_16S), names(genus_mean_proportions_WTS))  
set_WTS_only <- setdiff(names(genus_mean_proportions_WTS), names(genus_mean_proportions_16S))  

print(paste("Shared genus:", paste(intersection, collapse = ", ")))
print(paste("16S only:", paste(set_16S_only, collapse = ", ")))
print(paste("WTS only:", paste(set_WTS_only, collapse = ", ")))

set_16S <- names(genus_mean_proportions_16S)
set_WTS <- names(genus_mean_proportions_WTS)

venn_data <- list(
  Set1 = set_16S,
  Set2 = set_WTS
)

venn.plot <- venn.diagram(
  x = venn_data,
  category.names = c("16S", "WTS"),   
  filename = NULL,                     
  output = TRUE,                      
  fill = c("skyblue", "pink"),        
  alpha = 0.5,                        
  cex = 1.5,                           
  cat.cex = 1.5,                        
  cat.pos = c(-10, 10),                 
  cat.dist = c(0.05, 0.05),             
  lwd = 2,                             
  fontfamily = "sans",                 
  label.col = "transparent",           
  cat.col = c("blue", "red"),
  scaled = FALSE            
)

values <- genus_mean_proportions_WTS[set_WTS_only]
sorted_set_WTS_only <- set_WTS_only[order(values, decreasing = TRUE)]
rankings <- rank(-genus_mean_proportions_WTS)
sorted_set_WTS_only <- sapply(sorted_set_WTS_only, function(genus) {
  paste(genus, " (", rankings[genus],")",sep="")
})

values <- genus_mean_proportions_16S[set_16S_only]
sorted_set_16S_only <- set_16S_only[order(values, decreasing = TRUE)]
rankings <- rank(-genus_mean_proportions_16S)
sorted_set_16S_only <- sapply(sorted_set_16S_only, function(genus) {
  paste("(", rankings[genus],") ",genus,sep="")
})

values <- genus_mean_proportions_16S[setdiff(names(genus_mean_proportions_16S), set_16S_only)]
intersection_with_values <- intersection[order(values, decreasing = TRUE)]
rankings_16S <- rank(-genus_mean_proportions_16S)
rankings_WTS <- rank(-genus_mean_proportions_WTS)
intersection_with_values <- sapply(intersection_with_values, function(genus) {
  paste("(", rankings_16S[genus],") ", genus, " (", rankings_WTS[genus],")",sep="")
})

pdf("../Figures/WTS_16S_top10_venn_diagram_filtered.pdf", width = 9, height = 6)
grid.newpage()
grid.draw(venn.plot)

grid.text(paste(paste(intersection_with_values, collapse = "\n"), collapse = ", "), x = 0.5, y = 0.475, gp = gpar(fontsize = 12.5))
grid.text(paste(paste(sorted_set_16S_only, collapse = "\n"), collapse = ", "), x = 0.325, y = 0.475, gp = gpar(fontsize = 12.5), just = "right") 
grid.text(paste(paste(sorted_set_WTS_only, collapse = "\n"), collapse = ", "), x = 0.675, y = 0.475, gp = gpar(fontsize = 12.5), just = "left")
dev.off()
```



```{r CCC}
results_table <- data.frame(
  Normalization = character(),
  CCC_Genus_Median = numeric(),
  CCC_Sample_Richness = numeric(),
  CCC_Sample_Total_Reads = numeric(),
  Overlapped_Genera = numeric(),
  stringsAsFactors = FALSE
)
Norm_list <- c("Matched","Filtered","TSS","DRP","DV","TMM","CSS","GMPR","Wrench")

for (i in 1:9) {
  eval(parse(text  =  paste0("M_16S <- ",Norm_list[i],"_M_16S@otu_table")))
  eval(parse(text  =  paste0("M_WTS <- ",Norm_list[i],"_M_WTS@otu_table")))

  # CCC between genus-median
  median_16S <- apply(M_16S, 1, median)
  median_WTS <- apply(M_WTS, 1, median)
  ccc_genus_median <- CCC(median_16S, median_WTS)$rho.c$est

  # CCC between sample richness
  richness_16S <- apply(M_16S, 2, function(x) sum(x > 0))
  richness_WTS <- apply(M_WTS, 2, function(x) sum(x > 0))
  ccc_sample_richness <- CCC(richness_16S, richness_WTS)$rho.c$est

  # CCC between sample total reads
  total_reads_16S <- colSums(M_16S)
  total_reads_WTS <- colSums(M_WTS)
  ccc_sample_total_reads <- CCC(total_reads_16S, total_reads_WTS)$rho.c$est

  # top 10 genera overlapped #
  total_abundance_16S <- rowSums(M_16S)
  total_abundance_WTS <- rowSums(M_WTS)
  top10_16S <- names(sort(total_abundance_16S, decreasing = TRUE))[1:10]
  top10_WTS <- names(sort(total_abundance_WTS, decreasing = TRUE))[1:10]
  overlapped_genera <- length(intersect(top10_16S, top10_WTS))

  results_table <- rbind(results_table, data.frame(
    Version = Norm_list[i],
    CCC_Genus_Median = ccc_genus_median,
    CCC_Sample_Richness = ccc_sample_richness,
    CCC_Sample_Total_Reads = ccc_sample_total_reads,
    Overlapped_Genera = overlapped_genera
  ))
}
kable(results_table, 
      caption = "Results of CCC and Overlapped Genera for Different Normalization", 
      col.names = c("Version", "CCC Genus Median", "CCC Sample Richness", "CCC Sample Total Reads", "Overlapped Genera"),
      format = "markdown")
```