---
title: "DA method"
output: html_document
---

This code is used to generate ALDEx2 and ANCOMBC2 result. Note that for Aldex2 which requires counts, we round the normalized data if there are decimal points.

```{r load packages}
library(ALDEx2)
library(ANCOMBC)
library(dplyr)
```

The parameter "contrast" determines that subsequent code will generate plots using data derived from three different contrasts.

```{r load data}
 contrast = "CDUC vs nonIBD"
# contrast = "IBD vs nonIBD"

root <- ifelse(contrast == "IBD vs nonIBD", "IBDvsNonIBD","CDUCvsNonIBD")

Raw_M_16S = readRDS("Raw_M_16S.rds")
Raw_M_WTS = readRDS("Raw_M_WTS.rds")

Matched_M_16S = readRDS("Matched_M_16S.rds")
Matched_M_WTS = readRDS("Matched_M_WTS.rds")

Filtered_M_16S = readRDS("Filtered_M_16S.rds")
Filtered_M_WTS = readRDS("Filtered_M_WTS.rds")

TSS_M_16S = readRDS("TSS_M_16S.rds")
TSS_M_WTS = readRDS("TSS_M_WTS.rds")

DRP_M_16S = readRDS("DRP_M_16S.rds")
DRP_M_WTS = readRDS("DRP_M_WTS.rds")

DV_M_16S = readRDS("DV_M_16S.rds")
DV_M_WTS = readRDS("DV_M_WTS.rds")

TMM_M_16S = readRDS("TMM_M_16S.rds")
TMM_M_WTS = readRDS("TMM_M_WTS.rds")

CSS_M_16S = readRDS("CSS_M_16S.rds")
CSS_M_WTS = readRDS("CSS_M_WTS.rds")

GMPR_M_16S = readRDS("GMPR_M_16S.rds")
GMPR_M_WTS = readRDS("GMPR_M_WTS.rds")

Wrench_M_16S = readRDS("Wrench_M_16S.rds")
Wrench_M_WTS = readRDS("Wrench_M_WTS.rds")

data_list = c("Raw_M_16S","Raw_M_WTS",
              "Matched_M_16S","Matched_M_WTS",
              "Filtered_M_16S","Filtered_M_WTS",
              "TSS_M_16S","TSS_M_WTS",
              "DRP_M_16S","DRP_M_WTS",
              "DV_M_16S","DV_M_WTS",
              "TMM_M_16S","TMM_M_WTS",
              "CSS_M_16S","CSS_M_WTS",
              "GMPR_M_16S","GMPR_M_WTS",
              "Wrench_M_16S","Wrench_M_WTS")

if (contrast == "IBD vs nonIBD"){
  for (i in data_list){
    eval(parse(text = paste0(i,"@sam_data$diagnosis = recode(",i,"@sam_data$diagnosis, 'CD' = 'IBD', 'UC' = 'IBD')")))
  }
}
```

```{r define function for ALDEx2 DA}
DA_aldex <- function(ps){
  counts_tab <- round(ps@otu_table@.Data)
  covariates <- data.frame(diagnosis = ps@sam_data$diagnosis, 
                           bodysite = ps@sam_data$body_site,
                           age = ps@sam_data$consent_age,
                           row.names = colnames(counts_tab))
  covariates <- covariates[!is.na(covariates$age),]
  design <- model.matrix( ~ diagnosis + bodysite + age, covariates)
  results <- aldex(reads = counts_tab[,rownames(covariates)], 
                   conditions  =  design, 
                   mc.samples  =  128, 
                   test = "glm", 
                   effect = TRUE,
                   include.sample.summary  =  FALSE, 
                   verbose = T, 
                   denom = "all")
  return(results)
}
```

```{r generate Aldex2 and ANCOMBC DA result, warning=FALSE}
set.seed(111)
Datasets <- c("Matched","Filtered", "TSS","TMM","CSS","DRP","DV", "GMPR","Wrench")
for (type in c("16S","WTS")){
  for(dataset in Datasets){
      eval(parse(text  =  paste0("ps = ", dataset, "_M_", type)))
      #if(dataset == "Matched"){
      #  ps@tax_table = ps@tax_table[ps@tax_table != "Proteiniclasticum"]
      #  ps@otu_table = ps@otu_table[rownames(ps@otu_table) != "Proteiniclasticum",]
      #}
      aldex_res <- DA_aldex(ps)
      if(dataset == "Matched" & type =="WTS" & !("Proteiniclasticum" %in% rownames(aldex_res))){
        new_row <- data.frame(matrix(NA, ncol = ncol(aldex_res), nrow = 1)) 
        rownames(new_row) <- "Proteiniclasticum" 
        colnames(new_row) <- colnames(aldex_res)
        aldex_res <- rbind(aldex_res, new_row)
        aldex_res <- aldex_res[order(rownames(aldex_res)), ]
      }
      ancombc_res <- ancombc2(data = ps,
                              assay_name = "otu_table",
                              fix_formula = "body_site + diagnosis + consent_age",
                              p_adj_method = "none",
                              prv_cut = 0,
                              lib_cut = 0,
                              group = "diagnosis",
                              struc_zero = FALSE,
                              neg_lb = FALSE,
                              alpha = 0.01,
                              global = F,
                              tax_level = "Genus")
            
      aldex_res <- cbind(data.frame(taxon = rownames(aldex_res)),aldex_res)
     
      write.table(aldex_res, paste0("DAResults/", root, "/ALDEx2_", dataset, "_M_", type,".csv"),sep =",", row.names = F, col.names = T)
      write.table(ancombc_res$res, paste0("DAResults/", root, "/ANCOMBC_", dataset, "_M_", type,".csv"), sep =",", row.names = F, col.names = T)
  }
}
```

#The genus Proteiniclasticum appeared in the WTS data, but its occurrence was extremely low, appearing only once in a single sample with a read count of 2. As a result, aldex2 automatically removed this genus from the analysis, leading to a mismatch in the number of genera between the WTS results and the 16S results in the aldex2 matched data.
#The genus Proteiniclasticum also appeared in the 16S data, but its occurrence was similarly very low, appearing only once in 2 samples, with a read count of 3 in each.

For now, I am manually adding Proteiniclasticum to result of matched data-aldex2. 
